\textbf{Microbiota predict \emph{Clostridium difficile} severity in germ-free mice colonized with human feces}

### 2016-11-14 submission/Manuscript.Rmd

@@

**The microbiome predicts _C. difficile_ severity.** As _C. difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0. Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease. We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med). The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B). The populations that were most commonly found in mice that succumbed to _C. difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_. (need more OTUs here, or number that were used for model and then group by rank). The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes.

@@

b.	Figure 4: Random Forest predicts CDI severity, predictive OTUs
2.	Propensity for severe CDI is community-dependent and strain-independent
a.	Figure 5:  Infection of mice with different C. difficile strains.

### Discussion
•	Restate results,
•	caveats about mouse weights 
•	No donors were colonization resistant, discuss donor differences 
•	Discuss prediction methods and outcomes
•	Discuss potential mechanisms for interesting OTUs
•	Discuss different strain results
•	Future work blah blah

>>>>

**also want something in the intro about severity maybe, and asymptomatic colonization. or in discussion?**

@@

**The microbiome predicts _C. difficile_ infection severity.** As _C. difficile_ colonization levels can be predicted by the microbiome composition on day 0, we postulated that severity of disease could also be predicted from day 0. Using weight loss as a proxy for severity, we built a Random Forest classification model based on relative abundance to determine which community members were associated with either mild or severe disease. We optimized the model using the AUC-RF algorithm to generate a Reciever Operating Curve (ROC) with maximal Area Under the Curve (AUC) (cite Calle 2011), as used previously in our group to classify colon cancer outcomes (cite Niel 2015 gen med). The optimized model used X number of OTUs and identified several that predicted increased weightloss and subsequent disease severity (Fig 4A, 4B). The populations that were most commonly found in mice that succumbed to _C. difficile_ infection were members of the _Clostridiales_, _Bacilliales_, _Streptococcus_ (Lactobacilles?) and _Bacteriodes_. (need more OTUs here, or number that were used for model and then group by rank). The AUC was calculated using the Out-Of-Bag (OOB) error for each sample (Fig 4A). Cross-validation using 10-fold, leave-one-mouse-out and leave-one-cage-out resulted in AUCs not significantly different from the OOB (Fig 4A). This model reveals that the levels of X OTUs on day 0 can predict disease outcome in mice with human-associated microbiomes. (probably also going to want a supplemental table for predictive OTUs for this and Fig 3)

**Susceptibility to infection is community-dependent.** _C. difficile_ strains are known to vary in disease phenotypes based both on pathogenicity factors of the strain and host microbiome composition (cite someone). To test the susceptibility of human-derived microbiomes to _C. difficile_ infection, two donor-mouse microbiomes that were previously shown to be resistant to _C. difficile_ strain 431 induced weight loss (Fig 2B) (DA00396, DA00430) and one that was suscpetible to severe disease (DA00578) were independently challenged with one of three unique _C. difficile_ strains (table or reference for toxin prod, etc). As done previously, _C. difficile_ stool CFU was enumerated daily and mouse weight-loss tracked. Infection of mice with resistant communities with virulent strain 458 or the weaker strain 299 resulted in colonization to levels near that of 431 (Fig 5A) and these mice did not experience severe weight loss or show clinical signs of disease (Fig 5B). In contrast, infection of mice harboring microbiomes susceptible to severe disease (DA00578, Fig 1B) with strain 3

@@

2.	Propensity for severe CDI is community-dependent and strain-independent
a.	Figure 5:  Infection of mice with different C. difficile strains.

### Discussion
•	Restate results,
•	caveats about mouse weights 
•	No donors were colonization resistant, discuss donor differences 
severity, asymptomatic colonzation discussion? 
•	Discuss prediction methods and outcomes
•	Discuss potential mechanisms for interesting OTUs
•	Discuss different strain results
•	Future work blah blah

### 2016-11-14 submission/Manuscript.tex

Clostridium difficile is an opportunistic pathogen of the human lower
gastrointestinal tract. Disruption of the native microbial community of
the gut by antibiotics is the most common risk factor for development of
C. difficile infection (CDI) (1). C. difficile is a spore-forming
bacteria and can persist on abiotic surfaces and is not readily killed
by ethanol-based hand-sanitizers, putting hospital patients particularly
at risk. Indeed, \textasciitilde{}12\% of hospital acquired infections
in the United States are due to C. difficile and result in up to 15,000
deaths annually (2).

Murine models to study CDI typically rely on treating
conventionally-raised mice with antibiotics either in drinking water or
by injection to induce susceptibility (3, 4). This model provides a
convenient way to study C. difficile pathogenesis and virulence factors.
Numerous microbiome studies have been performed using this model to
determine the antibiotic classes (5), starting microbial community (6)
and metabolites (7) that impact development and severity of CDI. While
informative, these studies are somewhat removed from human disease
because they only examine mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of
studies of CDI, including assessment of species-specific interactions
between C. difficile and competing microbial community members (8),
analysis of nutrient restriction (9), in vivo transcriptomics of C.
difficile and examination of host immune response to CDI (10). Further,
CDI therapeutics such as antibiotics and fecal microbiota transplants
have been tested extensively in a gnotobiotic-piglet or
piglet-to-gnotobiotic-mouse model of disease (11), (12). Pigs have a
longer digestive tract with components more similar to humans than mice
and are typically infected by strains typical in human infection (13).
However, the murine and porcine microbiomes typically do not resemble
those of the human gut.

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with C. difficile none of the antibiotic-treated piglets
succumbed to infection, while some of the untreated human-colonized pigs
did (11). Further, germ-free mice colonized with human feces were bred
over several generations to create a cohort of mice with identical
human-derived microbiomes (14). These mice were subsequently treated
with a five-antibiotic cocktail to induce dysbiosis and then were
successfully colonized by C. difficile (14). While informative, these
studies were limited in their use of only one human donor as input
inoculum. In order to best understand the impact of C. difficile
pathogenesis on human disease, we must have a laboratory model that
allows for study of a variety of human-derived microbiomes.

To test the impact of individual human microbiomes on CDI, we colonized
germ-free mice with 16 different human stool donors. We then
characterized human-associated microbiome response to C. difficile
challenge. Additionally, the use of machine-learning models allowed us
to build a predictive model that classified ``at-risk'' microbiomes
prior to infection with C. difficile. These findings show that
human-associated microbiomes can be at risk for CDI even in the absence
of antibiotics and that study of mice colonized with human feces
provides a range of clinical outcomes.

@@

@@

\textbf{Results to be written} 1. Microbes present in the gut prior to
infection are predictive of C. difficile CFU and severity a. Figure 3:
Random forest to predict CFU, predictive OTUs b. Figure 4: Random Forest
predicts CDI severity, predictive OTUs 2. Propensity for severe CDI is
community-dependent and strain-independent a. Figure 5: Infection of
mice with different C. difficile strains.

@@

• Mice ULAM number • Donor stool ERIN IRB shit • Bacteria/plating •
Sequencing • Data analysis, code availability • Machine learning models

@@

\textbf{Figure 1. Germ-free mice inoculated with human feces as a model
for C. difficile infection.} A) Experimental design. Stool from 16
healthy, diarrheal and CDI patients were independently inoculated into
3-4 germ-free mice by oral gavage. 14 days later mice were orally
infected with 100 spores of C. difficile strain 431. Weight and stool
CFU were monitored for up to 10 days post infection. B) NMDS ordination
of donor stool communities prior to inoculating mice. Each point
represents one donor and are colored by clinical diagnosis. C) NDMS
ordination of the stool communities on day 0. Each symbol represents one
mouse and is colored by donor. Circles represent mice that experienced
mild disease and triangles represent those that suffered severe disease.

@@

@@

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

>>>>

\emph{Clostridium difficile} is an opportunistic pathogen of the human
lower gastrointestinal tract. \emph{C. difficile} forms spores that can
persist on abiotic surfaces and are not readily killed by ethanol-based
hand-sanitizers, putting hospital patients particularly at risk. Indeed,
\textasciitilde{}12\% of hospital acquired infections in the United
States are due to \emph{C. difficile} and result in up to 15,000 deaths
annually (2). Disruption of the native microbial community is the most
common risk factor for development of \emph{C. difficile} infection
(CDI) (1). Antibiotic use and inflammatory bowel diseases are associated
with loss of colonization resistance to \emph{C. difficile} through the
loss of potentially protective bacterial families such as
\emph{Barnesiella} and \emph{Lachnospiraceae} in both mouse models and
human association experiments (cite alyx 2014 and vincent 2013). The
composition of the community is clearly important for the acquisition,
resistance and treatment of \emph{C. difficile} infection, as giving
patients a healthy fecal microbiome transplant is the most effective
treatment for this disease (cite Anna 2014). The precise mechanisms of
colonization resistance and restoration of a healthy community are yet
to be discovered.

Murine models to study CDI typically rely on treating
conventionally-raised mice with antibiotics either in drinking water or
by injection to induce susceptibility (3, 4). This model provides a
convenient way to study \emph{C. difficile} pathogenesis and virulence
factors. Numerous microbiome studies have been performed using this
model to determine the antibiotic classes (5), starting microbial
community (6) and metabolites (7) that impact development and severity
of \emph{C. difficile} infection. While informative, these studies are
somewhat removed from human disease because they only examine
mouse-associated microbial communities.

Gnotobiotic or germ-free mouse models have been used for a range of
studies of CDI, including assessment of species-specific interactions
between \emph{C. difficile} and competing microbial community members
(8), analysis of nutrient restriction (9), \emph{in vivo}
transcriptomics of \emph{C. difficile} and examination of host immune
response to CDI (10). Further, CDI therapeutics such as antibiotics and
fecal microbiota transplants have been tested extensively in a
gnotobiotic-piglet or piglet-to-gnotobiotic-mouse model of disease (11),
(12). Pigs have a longer digestive tract with components more similar to
humans than mice and are typically infected by strains typical in human
infection (13). However, the murine and porcine microbiomes typically do
not resemble those of the human gut.

The power of the gnotobiotic models to study CDI has been further
realized by first inoculating germ-free mice and piglets with human
stool microbes. In one study, germ-free piglets were acutely colonized
with human feces for one week and then treated with tigecycline. After
challenge with C. difficile none of the antibiotic-treated piglets
succumbed to infection, while some of the untreated human-colonized pigs
did (11). In another study, germ-free mice colonized with human feces
were bred over several generations to create a cohort of mice with
identical human-derived microbiomes (14). These mice were subsequently
treated with a five-antibiotic cocktail to induce dysbiosis and then
were successfully colonized by \emph{C. difficile} (14). While
informative, these studies were limited in their use of only one human
donor as input inoculum. In order to best understand the impact of
\emph{C. difficile} pathogenesis on human disease, we must have a
laboratory model that allows for study of a variety of human-derived
microbiomes.

Here we designed a laboratory model that allows for studying and
modeling microbial coumminity interactions with \emph{C. difficile}
infection in mice with human derived microbiomes. To test the impact of
individual human microbiomes on CDI, we colonized germ-free mice with 16
different human stool donors. We then characterized human-associated
microbiome response to C. difficile challenge. Additionally, the use of
machine-learning models allowed us to build a predictive model that
classified ``at-risk'' microbiomes prior to infection with C. difficile.
These findings show that human-associated microbiomes can be at risk for
CDI even in the absence of antibiotics and that study of mice colonized
with human feces provides a range of clinical outcomes.

@@

\textbf{need to put in a line to reference supplemental table with RA in
it}

@@

\textbf{Microbes present in the gut prior to infection are predictive of
\emph{C. difficile} colonization levels} Previous work in our group has
demonstrated that the microbes present in the mouse microbiome prior to
\emph{C. difficile} challenge can predict future colonization levels
(cite Alyx). To determine if the human-associated microbiome in the
mouse could similarly predict \emph{C. difficile} colonization we
employed a Random Forest machine learning algorithm similar to the one
used in previous studies. Relative abundance of OTUs with greater than
1\% abundance at day 0 were used as input to the model. We found that
the model explained X\% of the variance in colonization levels of
\emph{C. difficile} (Fig 3A). By refining the model to include only the
top X predictive OTUs, the model explained X\% of the variance. Partial
dependency plots of the top 6 OTUs predictive of \emph{C. difficile}
colonization reveal that colonization is positively associated with
\emph{Bacteriodetes}, \emph{Parabacteriodes} and
\emph{Porphyromonadaceae} (really?) (Fig 3B). Colonization was
negatively associated with the presence of \emph{Lachnospiraceae} (Fig
3B) though none of the human-associated mouse communities were
completely resistant to colonization by \emph{C. difficile} (Fig 2A,
3B). These results confirm that as in mouse microbiome studies, the
human-associated gut community prior to challenge is predictive of
\emph{C. difficile} colonization levels.

\textbf{The microbiome predicts \emph{C. difficile} severity.}

\textbf{Results to be written}

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Figure 4: Random Forest predicts CDI severity, predictive OTUs
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Propensity for severe CDI is community-dependent and
  strain-independent
\end{enumerate}

\begin{enumerate}
\def\labelenumi{\alph{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Figure 5: Infection of mice with different C. difficile strains.
\end{enumerate}

@@

\textbf{Mice} The experiments in this study used 6-8 week-old male and
female gnotobiotic C57/BL6 mice. The mice were born and maintainted in a
completely sterile environment in the Germ Free Mouse Facility at the
University of Michigan. The University Committee on Use and Care of
Animals approved all protocols and experiments conducted in this study.

\textbf{Human samples} Donor stool samples were chosen from a sample
bank collected from patients from October 2010 to November 2012 at the
University of Michigan. Sample collection methods and study was approved
by the Institutional Review Board at the University of Michigan. Stool
samples from University of Michigan Health System patients were
submitted to the microbiology lab for \emph{C. difficile} testing.
\emph{C. difficile} presence was measured using WHATEVER- Cdiff quik
check? toxin AB?. Samples from these patients were stored in Cary-Blair
medium at -80C. All patients consented to participation in the study.
Donor stool for this study was selected from one patient that tested
positive for \emph{C. difficile} infection and four patients that
presented with diarrhea that tested negative for \emph{C. difficile}.
Healthy donor stool was collected from volunteer donors living in the
Ann Arbor, MI area.

• Bacteria/plating • Sequencing • Data analysis, code availability •
Machine learning models

@@

\textbf{Figure 1. Germ-free mice inoculated with human feces as a model
for \emph{C. difficile infection.}} A) Experimental design. Stool from
16 healthy, diarrheal and CDI patients were independently inoculated
into 3-4 germ-free mice by oral gavage. 14 days later mice were orally
infected with 100 spores of C. difficile strain 431. Weight and stool
CFU were monitored for up to 10 days post infection. B) NMDS ordination
of donor stool communities prior to inoculating mice. Each point
represents one donor and are colored by clinical diagnosis. C) NDMS
ordination of the stool communities on day 0. Each symbol represents one
mouse and is colored by donor. Circles represent mice that experienced
mild disease and triangles represent those that suffered severe disease.

@@

\textbf{Figure 5. Infection of mice with different \emph{C. difficile
strains}.} 3 strains of \emph{C. difficile} were used to infect mice
colonized with susceptible (DA00578) or resistant (DA00369, DA00430)
human donor stool. A) \emph{C. difficile} stool CFU was enumerated over
10 days. B) Percent weight loss was calculated each day for each mouse.
In both plots, each mouse is a point and lines represent the mean of
each cage.

\paragraph{Supplement}\label{supplement}

\textbf{Table S1: Mouse day 0 communities by donor genera (avg + stdev
of cage)}

@@

\subsection{References}\label{references}
